---
title: ST503 Section 601 Homework 6
author: Charles Lane
format:
  pdf:
    mainfont: "Times New Roman"
---

Installation of necessary R packages:

```{r}
#| warning: false
library('tidyverse')
library('faraway')
library('lmtest')
library('rpart')
library('emmeans')
```
****
{{< pagebreak >}} 

## 1) Exercise 15.6 on page 233-234 of LMR
Data on the butterfat content of milk from Canadian cows of five different breeds can be found in the 'butterfat' dataset. Consider only mature cows.

### a) Plot the data and interpret what you see.
```{r}
butterfat$Age <- as.factor(butterfat$Age)
butterfat$Breed <- as.factor(butterfat$Breed)

b.p <- ggplot(data = subset(butterfat, Age == "Mature"), 
              aes(x = Breed, y = Butterfat))

b.p + 
  geom_boxplot() + 
  labs(title = "Butterfat vs Breed for Mature Cows")
```
The plot above appears to show an effect on Butterfat content from certain breeds, namely Jersey and Guernsey cows in the positive direction, and Holstein-Fresian cows in the negative direction.

### b) Test for a difference between the breeds.
```{r}
new_bf <- subset(butterfat, Age == "Mature")

lm.bf <- lm(Butterfat ~ Breed, data = new_bf)
bf1 <- aov(Butterfat ~ Breed, data = new_bf)
summary(bf1)
```
The results of the Analysis of Variance model show that there is a statistically significant difference between the various breeds, but what it does not communicate is which breed pair-wise combinations are statistically significant.


### c) Check the diagnostics for your model.
```{r}
qqnorm(residuals(lm.bf))
qqline(residuals(lm.bf))
plot(jitter(fitted(lm.bf)), 
     residuals(lm.bf), 
     xlab = "Fitted", 
     ylab = "Residuals")
abline(h = 0)
```
Both the normality assumption (Q-Q Plot) and the homogeneous error variance (residuals vs Fitted) assumptions appear to be in question. Further testing is below:
```{r}
#Normality of residuals assumption test using Shapiro-Wilk test
shapiro.test(residuals(lm.bf))
```
Based on the Shapiro-Wilk result, which is just above the $\alpha$ = 0.05 significant test, the dataset does conform with normality assumptions.

```{r}
#Constant Error Variance assumption test using Breusch-Pagan
bptest(lm.bf)
```
The Breusch-Pagan test results do not indicate a rejection of the null hypothesis, and thus not indicating the homogeneous error variance assumption is violated for this dataset.

### d) Produce a plot to check for differences in butterfat between breeds. Which pairwise differences are not statistically significant?
```{r}
plot(TukeyHSD((bf1)))
TukeyHSD(bf1)
```
Based on the plot above, and the Tukey comparison of all pairwise combinations, all pairwise comparisons are significantly different except for 2:

1. Canadian-Ayrshire
2. Jersey-Guernsey

****
{{< pagebreak >}} 

## 2) A study was conducted to compare 3 treatments for patients with HIV. The response was a measure of change in CD4+ counts. There were n = 90 patients in each of the 3 treatment conditions. The sample means and MSE are given below, note that positive responses implying increasing counts (good) and negative responses imply decreasing counts (bad).

$$\bar{Y}_{1\cdot} = 12.2$$
$$\bar{Y}_{2\cdot} = 5.1$$
$$\bar{Y}_{3\cdot} = -0.3$$
$$MSE = 400$$

### a) Fill in an ANOVA table. Give the conclusion based on the F-test.
```{r}
n_i <- 90
I <- 3
df_r <- I-1
df_e <- (n_i*I) - I
df_t <- (n_i*I) - 1
y_1 <- 12.2
y_2 <- 5.1
y_3 <- -0.3
y_g <- sum(y_1, y_2, y_3)/3
MSE <- 400

SSR <- n_i*(y_1 - y_g)^2 + n_i*(y_2 - y_g)^2 + n_i*(y_3 - y_g)^2
MSR <- SSR/(df_r)
SSE <- MSE*df_e
SST <- SSR + SSE
F_stat <- MSR/MSE
print(paste("regression degrees of freedom = ", df_r))
print(paste("Error degrees of freedom = ", df_e))
print(paste("Total degrees of freedom = ", df_t))
print(paste("Regression Sum of Squares = ", SSR))
print(paste("Error Sum of Squares = ", SSE))
print(paste("Total Sum of Squares = ",SST))
print(paste("MSR = ",MSR))
print(paste("MSE = ",MSE))
print(paste("F Observed Statistic = ",F_stat))
```

| Source  |   df   | Sum of Squares | Mean Squares | F-Statistic |
|---------|--------|----------------|--------------|-------------|
|   Reg   |    2   |     7074.6     |    3537.3    |   8.84325   |
|  Error  |   267  |     106800     |      400     |             |
|  Total  |   269  |    113874.6    |              |             |

```{r}
F_critical <- qf(0.95, df_r, df_e) # significance level of 0.05
HIV_p <- pf(F_stat, df_r, df_e, lower.tail = FALSE)
print(paste("Critical F-statistic = ",F_critical))
print(paste("p-value = ",HIV_p))
```

Based on the comparison between the critical & observed F-statistics, and the resulting p-value, the null hypothesis may be rejected at $\alpha$ = 0.05. The null hypothesis states there is no difference in group means. We may assume there are significant differences between at least some portion of group means for this dataset.

### b) Compute simultaneous 95% CIs for difference between all pairs of treatments, based on Tukey’s, and Bonferroni’s methods.
```{r}
# Tukey's
q_crit <- qtukey(p = 0.95, nmeans = 3, df = df_e)/sqrt(2)
hiv_se <- sqrt(MSE)*sqrt(2/n_i)
t.comp12 <- c((y_1 - y_2) - q_crit*hiv_se, 
              (y_1 - y_2) + q_crit*hiv_se)
t.comp23 <- c((y_2 - y_3) - q_crit*hiv_se, 
              (y_2 - y_3) + q_crit*hiv_se)
t.comp13 <- c((y_1 - y_3) - q_crit*hiv_se, 
              (y_1 - y_3) + q_crit*hiv_se)
print(paste("Tukey's CI for Treatments 1 vs 2 = ",t.comp12))
print(paste("Tukey's CI for Treatments 2 vs 3 = ",t.comp23))
print(paste("Tukey's CI for Treatments 1 vs 3 = ",t.comp13))
```
For the Tukey's calculation, the 95% CI for pairwise comparisons 1 vs 2 and 1 vs 3 do not include 0, indicating there is significant difference between the treatments 1 vs 2 & 2 vs 3.

```{r}
# Bonferroni - 3 Levels of treatment, 3 possible pairwise comparisons
J = 3
bonf_alpha <- 1 - ((1 - 0.05)^(1/J))
t_crit <- qt(p = 1 - bonf_alpha/2, df_e)
b.comp12 <- c((y_1 - y_2) - t_crit*hiv_se, 
              (y_1 - y_2) + t_crit*hiv_se)
b.comp23 <- c((y_2 - y_3) - t_crit*hiv_se, 
              (y_2 - y_3) + t_crit*hiv_se)
b.comp13 <- c((y_1 - y_3) - t_crit*hiv_se, 
              (y_1 - y_3) + t_crit*hiv_se)
print(paste("Bonferroni CI for Treatments 1 vs 2 = ",b.comp12))
print(paste("Bonferroni CI for Treatments 2 vs 3 = ",b.comp23))
print(paste("Bonferroni CI for Treatments 1 vs 3 = ",b.comp13))
```
For the Bonferroni calculation, the 95% CI for pairwise comparison 1 vs 3 does not include 0, indicating there is significant difference between these treatments 1 & 3 only.

### c) Carry out a test of contrast comparing the first two treatment with the last treatment.
```{r}
h_contrasts <- c(0.5, 0.5, -1)
h_means <- c(y_1, y_2, y_3)
cont_est <- sum(h_contrasts*h_means)
cont_se <- sqrt(MSE * sum((h_contrasts^2) / n_i))
t.stat <- cont_est/cont_se
cont_h_p <- 2 * pt(abs(t.stat), df = df_e, lower.tail = FALSE)
print(paste("Contrast Estimate = ",cont_est))
print(paste("T-Statistic = ",t.stat))
print(paste("p-value",cont_h_p))
```

With a p-value < $\alpha$ = 0.05, the null hypothesis may be rejected. This implies the group mean for the average of treatments 1 & 2 is significantly different from the group mean for treatment 3.

****
{{< pagebreak >}} 

## 3) Exercise 5.1 on page 99 of ELMR
The dataset 'discoveries' lists the numbers of "great" inventions and scientific discoveries in each year from 1860 to 1959.
### a) Plot the discoveries over time and comment on the trend, if any.
```{r}
plot(discoveries, ylab = "Number of Important Discoveries", 
     las = 1)
```
The trend appears to have a sharp increase in number of "great" discoveries in the 1880's, followed by a volatile, but on average reduction in number of great discoveries over time.

### b) Fit a Poisson response model with a constant term. Now compute the mean number of discoveries per year. What is the relationship between this mean and the coefficient seen in the model?
```{r}
df.d.p <- data.frame(discoveries)
df.d.p <- df.d.p %>% mutate(year = 1860:1959)

lm.d.p.null <- glm(discoveries ~ 1, family = "poisson", df.d.p)
summary(lm.d.p.null)
```

```{r}
mean.d.p <- mean(discoveries)

print(paste("Model Intercept (beta_0):", lm.d.p.null$coefficients))
print(paste("Mean Discoveries:", mean.d.p))
```
The coefficient & mean are related through a log-link function. Meaning, the coefficient is the log of the mean.

### c) Use the deviance from the model to check whether the model fits the data. What does this say about whether the rate of discoveries is constant over time?
```{r}
chi_crit <- qchisq(0.95, df = 99, lower.tail = FALSE)
pval.d.p <- 1 - pchisq(lm.d.p$deviance, df = 99)

print(paste("Null Deviance: ", lm.d.p$deviance))
print(paste("P-value: ", pval.d.p))
```

with a p-value less than $\alpha$ = 0.05, the deviance may be considered significant and the null hypothesis rejected. There is a dependency on year, or at least a difference across years that indicate discoveries were not constant across years.

### d) Make a table of how many years had zero, one, two, three, etc. discoveries. Collapse eight or more into a single category. Under an appropriate Poisson distribution, calculate the expected number of years with each number of discoveries. Plot the observed against the expected using a different plotting character to denote the number of discoveries. How well do they agree?
```{r}
tbl.d.p <- as.data.frame(table(factor(discoveries, 
                                      levels = 0:max(discoveries))))
colnames(tbl.d.p) <- c("n_discoveries", "years")
tbl.d.p$n_discoveries <- as.numeric(as.character(tbl.d.p$n_discoveries))

tbl_0_7 <- tbl.d.p$years[1:8]
tbl_8_plus <- sum(tbl.d.p$years[9:nrow(tbl.d.p)])
full.tbl <- c(tbl_0_7, tbl_8_plus)

lambda <- mean(discoveries)
n_years <- length(discoveries)

# Probabilities for 0-7, then 1 - P(X <= 7) for the "8+" category
exp_probs <- c(dpois(0:7, lambda), 1 - ppois(7, lambda))
expected <- exp_probs * n_years

plot(full.tbl, expected, 
     pch = c(as.character(0:7), "+"), 
     col = "blue",
     main = "Observed vs. Expected Years per Discovery Count",
     xlab = "Observed Number of Years", 
     ylab = "Expected Number of Years (Poisson)",
     xlim = c(0, 25), ylim = c(0, 25))
abline(0, 1, lty = 2, col = "red")
```
The Observed & Expected (based on a poisson model) number of discoveries, do follow a relatively consistent linear trend. While the observed counts are not perfectly in-line with the model, they do generally follow the linear path (linear as defined by the plot above).

```{r}
comparison <- data.frame(
  Discoveries = c(0:7, "8+"),
  Observed = full.tbl,
  Expected = round(expected, 2)
)
comparison
```


### e) Use the Pearson's Chi-squared test to check whether the observed numbers are consistent with the expected numbers. Interpret the result.
```{r}
#| warning: false
chi_test <- chisq.test(full.tbl, p = exp_probs)

stat <- chi_test$statistic
p_val_adj <- 1 - pchisq(stat, df = 7)

cat("Chi-squared Statistic:", stat, "\n")
cat("Adjusted P-value (df=7):", p_val_adj, "\n")
```

Given the p-value is less than $\alpha$ = 0.05, we may reject the null hypothesis and state that the observed values are NOT consistent with those expected from a poisson distribution. There are likely external factors or influential discoveries that impact future influential discoveries. Neither of these factors are accommodated in a Poisson model.

### f) Fit a Poisson response model that is quadratic in the year. Test for the significance of the quadratic term. What does this say about the presence of a trend in discovery?
```{r}
quad_exp <- glm(discoveries ~ year + I(year^2), 
                family = "poisson", 
                data = df.d.p)
summary(quad_exp)

pois.lm <- glm(discoveries ~ year, 
               family = "poisson", 
               data = df.d.p)

#Test explicitly by comparing models w/ and w/o the quadratic term
anova(pois.lm, quad_exp)
```

The p-value above is < $\alpha$ = 0.05, which implies we may reject the null hypothesis that the reduced model more adequately fits the data as compared to the full model. The quadratic term should be included in the model. Including a quadratic term indicates there is some amount of curvature, or possibly parabolic form to a model that better fits the rate of discoveries by year.

### g) Compute the predicted number of discoveries each year and show these predications as a line drawn over the data. Comment on what you see.
```{r}
df_disc <- data.frame(
  count = as.numeric(discoveries),
  year = 1860:1959
)

mod_quad <- glm(count ~ year + I(year^2), 
                family = "poisson", 
                data = df_disc)

df_disc$pred <- predict(mod_quad, type = "response")

ggplot(df_disc, aes(x = year, y = count)) +
  geom_point() +
  geom_line(aes(y = pred), color = "blue") +
  labs(title = "Poisson Regression: Quadratic Trend of Discoveries",
       x = "Year",
       y = "Number of Discoveries") +
  theme_minimal()
```
While there are still large "jumps" in number of discoveries in the decades of 1880's, 1910's, & 1920's, the quadratic model does capture the increase and gradual decrease in number of significant discoveries over time. i.e. the model captures the general shape, even if there are still potentially influential years.
