---
title: ST503 Section 601 Homework 6
author: Charles Lane
format:
  pdf:
    mainfont: "Times New Roman"
---

Installation of necessary R packages:

```{r}
#| warning: false
library('tidyverse')
library('faraway')
library('lmtest')
library('rpart')
library('emmeans')
```
****
{{< pagebreak >}} 

## 1) Exercise 15.6 on page 233-234 of LMR
Data on the butterfat content of milk from Canadian cows of five different breeds can be found in the 'butterfat' dataset. Consider only mature cows.

### a) Plot the data and interpret what you see.
```{r}
butterfat$Age <- as.factor(butterfat$Age)
butterfat$Breed <- as.factor(butterfat$Breed)

b.p <- ggplot(data = subset(butterfat, Age == "Mature"), 
              aes(x = Breed, y = Butterfat))

b.p + 
  geom_boxplot() + 
  labs(title = "Butterfat vs Breed for Mature Cows")
```
The plot above appears to show an effect on Butterfat content from certain breeds, namely Jersey and Guernsey cows in the positive direction, and Holstein-Fresian cows in the negative direction.

### b) Test for a difference between the breeds.
```{r}
new_bf <- subset(butterfat, Age == "Mature")

lm.bf <- lm(Butterfat ~ Breed, data = new_bf)
bf1 <- aov(Butterfat ~ Breed, data = new_bf)
summary(bf1)
```
The results of the Analysis of Variance model show that there is a statistically significant difference between the various breeds, but what it does not communicate is which breed pair-wise combinations are statistically significant.


### c) Check the diagnostics for your model.
```{r}
qqnorm(residuals(lm.bf))
qqline(residuals(lm.bf))
plot(jitter(fitted(lm.bf)), 
     residuals(lm.bf), 
     xlab = "Fitted", 
     ylab = "Residuals")
abline(h = 0)
```
Both the normality assumption (Q-Q Plot) and the homogeneous error variance (residuals vs Fitted) assumptions appear to be in question. Further testing is below:
```{r}
#Normality of residuals assumption test using Shapiro-Wilk test
shapiro.test(residuals(lm.bf))
```
Based on the Shapiro-Wilk result, which is just above the $\alpha$ = 0.05 significant test, the dataset does conform with normality assumptions.

```{r}
#Constant Error Variance assumption test using Breusch-Pagan
bptest(lm.bf)
```
The Breusch-Pagan test results do not indicate a rejection of the null hypothesis, and thus not indicating the homogeneous error variance assumption is violated for this dataset.

### d) Produce a plot to check for differences in butterfat between breeds. Which pairwise differences are not statistically significant?
```{r}
plot(tukey.bf)
TukeyHSD(bf1)
```
Based on the plot above, and the Tukey comparison of all pairwise combinations, all pairwise comparisons are significantly different except for 2:

1. Canadian-Ayrshire
2. Jersey-Guernsey

## 2) A study was conducted to compare 3 treatments for patients with HIV. The response was a measure of change in CD4+ counts. There were n = 90 patients in each of the 3 treatment conditions. The sample means and MSE are given below, note that positive responses implying increasing counts (good) and negative responses imply decreasing counts (bad).

$$\bar{Y}_{1\cdot} = 12.2$$
$$\bar{Y}_{2\cdot} = 5.1$$
$$\bar{Y}_{3\cdot} = -0.3$$
$$MSE = 400$$

### a) Fill in an ANOVA table. Give the conclusion based on the F-test.
```{r}
n_i <- 90
I <- 3
df_r <- I-1
df_e <- (n_i*I) - I
df_t <- (n_i*I) - 1
y_1 <- 12.2
y_2 <- 5.1
y_3 <- -0.3
y_g <- sum(y_1, y_2, y_3)/3
MSE <- 400

SSR <- n_i*(y_1 - y_g)^2 + n_i*(y_2 - y_g)^2 + n_i*(y_3 - y_g)^2
MSR <- SSR/(df_r)
SSE <- MSE*df_e
SST <- SSR + SSE
F_stat <- MSR/MSE
print(paste("regression degrees of freedom = ", df_r))
print(paste("Error degrees of freedom = ", df_e))
print(paste("Total degrees of freedom = ", df_t))
print(paste("Regression Sum of Squares = ", SSR))
print(paste("Error Sum of Squares = ", SSE))
print(paste("Total Sum of Squares = ",SST))
print(paste("MSR = ",MSR))
print(paste("MSE = ",MSE))
print(paste("F Observed Statistic = ",F_stat))
```

| Source  |   df   | Sum of Squares | Mean Squares | F-Statistic |
|---------|--------|----------------|--------------|-------------|
|   Reg   |    2   |     7074.6     |    3537.3    |   8.84325   |
|  Error  |   267  |     106800     |      400     |             |
|  Total  |   269  |    113874.6    |              |             |

```{r}
F_critical <- qf(0.95, df_r, df_e) # significance level of 0.05
HIV_p <- pf(F_stat, df_r, df_e, lower.tail = FALSE)
print(paste("Critical F-statistic = ",F_critical))
print(paste("p-value = ",HIV_p))
```

Based on the comparison between the critical & observed F-statistics, and the resulting p-value, the null hypothesis may be rejected at $\alpha$ = 0.05. The null hypothesis states there is no difference in group means. We may assume there are significant differences between at least some portion of group means for this dataset.

### b) Compute simultaneous 95% CIs for difference between all pairs of treatments, based on Tukey’s, and Bonferroni’s methods.
```{r}
# Tukey's
q_crit <- qtukey(p = 0.95, nmeans = 3, df = df_e)/sqrt(2)
t.comp12 <- c((y_1 - y_2) - q_crit*hiv_se, 
              (y_1 - y_2) + q_crit*hiv_se)
t.comp23 <- c((y_2 - y_3) - q_crit*hiv_se, 
              (y_2 - y_3) + q_crit*hiv_se)
t.comp13 <- c((y_1 - y_3) - q_crit*hiv_se, 
              (y_1 - y_3) + q_crit*hiv_se)
print(paste("Tukey's CI for Treatments 1 vs 2 = ",t.comp12))
print(paste("Tukey's CI for Treatments 2 vs 3 = ",t.comp23))
print(paste("Tukey's CI for Treatments 1 vs 3 = ",t.comp13))
```
For the Tukey's calculation, the 95% CI for pairwise comparisons 1 vs 2 and 1 vs 3 do not include 0, indicating there is significant difference between the treatments 1 vs 2 & 2 vs 3.

```{r}
# Bonferroni - 3 Levels of treatment, 3 possible pairwise comparisons
J = 3
bonf_alpha <- 1 - ((1 - 0.05)^(1/J))
hiv_se <- sqrt(MSE)*sqrt(2/n_i)
t_crit <- qt(p = 1 - bonf_alpha/2, df_e)
b.comp12 <- c((y_1 - y_2) - t_crit*hiv_se, 
              (y_1 - y_2) + t_crit*hiv_se)
b.comp23 <- c((y_2 - y_3) - t_crit*hiv_se, 
              (y_2 - y_3) + t_crit*hiv_se)
b.comp13 <- c((y_1 - y_3) - t_crit*hiv_se, 
              (y_1 - y_3) + t_crit*hiv_se)
print(paste("Bonferroni CI for Treatments 1 vs 2 = ",b.comp12))
print(paste("Bonferroni CI for Treatments 2 vs 3 = ",b.comp23))
print(paste("Bonferroni CI for Treatments 1 vs 3 = ",b.comp13))
```
For the Bonferroni calculation, the 95% CI for pairwise comparison 1 vs 3 does not include 0, indicating there is significant difference between these treatments 1 & 3 only.

### c) Carry out a test of contrast comparing the first two treatment with the last treatment.
```{r}
h_contrasts <- c(0.5, 0.5, -1)
h_means <- c(y_1, y_2, y_3)
cont_est <- sum(h_contrasts*h_means)
cont_se <- sqrt(MSE * sum((h_contrasts^2) / n_i))
t.stat <- cont_est/cont_se
cont_h_p <- 2 * pt(abs(t.stat), df = df_e, lower.tail = FALSE)
print(paste("Contrast Estimate = ",cont_est))
print(paste("T-Statistic = ",t.stat))
print(paste("p-value",cont_h_p))
```

With a p-value < $\alpha$ = 0.05, the null hypothesis may be rejected. This implies the group mean for the average of treatments 1 & 2 is significantly different from the group mean for treatment 3.

## 3)